---
title: "Clinical"
author: "Qirui Zhang"
date: "2025-04-07"
output: 
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

In total, **160 analyses** were conducted across **20 tumor types**, involving the following **8 distinct phenotypes**:

- Aneuploidy Score  
- Fraction Genome Altered  
- Months of Disease-specific Survival  
- Tumor Mutation Burden (TMB nonsynonymous)  
- Diagnosis Age  
- Progression-Free Survival Months  
- EUR (derived from the categorical variable Genetic Ancestry Label by coding European ancestry and European ancestry mix as 1, all others as 0)  
- AFR (derived from the categorical variable Genetic Ancestry Label by coding African ancestry and African ancestry mix as 1, all others as 0)

Since the original **Genetic Ancestry Label** phenotype was categorical, it was transformed into two binary variables (**EUR** and **AFR**) as described above, resulting in a total of 8 phenotypes for analysis. All **160 analyses** were successfully completed without errors.

# Extracting and Summarizing Analysis Results
This code segment scans through analysis output directories, identifies relevant result files (*_resdd.Rd), extracts corresponding tumor types, immune phenotypes, and analysis modes(sig/reg).
```{r}
get_files <- function(outputdir, tumors = NULL, phenotypes = NULL) {
  files <- list.files(
    path = outputdir,
    pattern = "_resdd\\.Rd$",
    recursive = TRUE,
    full.names = TRUE
  )
  
  matching_files_txt <- sub("\\.Rd$", ".txt", files)
  
  analysis_mode <- ifelse(grepl("_sig_", basename(files)), "sig", "reg")

  inner_folder_names      <- basename(dirname(files))            # e.g. "Aneuploidy.Score_BLCA_mutations"
  phenotype_folder_names  <- basename(dirname(dirname(files)))   # e.g. "Aneuploidy.Score"
  
  tumor_names <- mapply(function(inner_name, phen) {
    pattern_front <- paste0("^", phen, "_")
    tmp <- sub(pattern_front, "", inner_name)
    tmp <- sub("_mutations$", "", tmp)
    return(tmp)
  }, inner_folder_names, phenotype_folder_names)
  
  phenos_all <- phenotype_folder_names
  
  filedf <- data.frame(
    tumor      = tumor_names,
    phenotype  = phenos_all,
    mode       = analysis_mode,   
    file       = files,
    filetxt    = matching_files_txt,
    stringsAsFactors = FALSE
  )
  
  if (!is.null(tumors)) {
    filedf <- filedf[filedf$tumor %in% tumors, ]
  }
  if (!is.null(phenotypes)) {
    filedf <- filedf[filedf$phenotype %in% phenotypes, ]
  }

  rownames(filedf) <- NULL
  return(filedf)
}

filedf <- get_files(
  outputdir = "/dartfs/rc/lab/S/Szhao/qiruiz/diffdriver/temp/output/clinical"
)

head(filedf)
```

# Summary Table of Significant Differential Genes
The provided function extracts and compiles results from analyses, filtering specifically for significant differential genes based on a threshold (**`dd.fdr < 0.1`**). 
```{r}
get_diff_table <- function(filedf){
  pheno_all <- unique(filedf$phenotype)
  numlist <- list()
  
  for (p in pheno_all){
    p_txtfiles <- filedf[filedf$phenotype == p, ]
    numlist[[p]] <- list()
    
    for (t in seq_len(nrow(p_txtfiles))){
      txtf  <- p_txtfiles[t, "filetxt"]
      rdf   <- p_txtfiles[t, "file"] 
      tumor <- p_txtfiles[t, "tumor"] 
      mode  <- p_txtfiles[t, "mode"]     # reg / sig
      
      env <- new.env()
      load(rdf, envir = env)
      res_rdata <- env$res
      
      res <- read.table(txtf, header = TRUE)
      res$gene <- row.names(res)
      res$mode <- mode
      
      res$alpha <- sapply(res$gene, function(gene){
          res_rdata[[gene]][["dd"]][["res.alt"]]$alpha[2]
      })
      
      sig_res <- res[res$dd.fdr < 0.1, ]
      if (nrow(sig_res) > 0){
        numlist[[p]][[ paste0(tumor, "_", mode) ]] <- sig_res
      }
    }
    
    if (length(numlist[[p]]) == 0){
      numlist[[p]] <- NULL
    }
  }
  
  combined_df <- do.call(
    rbind,
    lapply(names(numlist), function(pheno) {
      numlist.pheno <- numlist[[pheno]]
      df.pheno <- do.call(rbind, lapply(names(numlist.pheno), function(tname) {
        df <- numlist.pheno[[tname]]

        df$tumor <- sub("_.*", "", tname) 
        df$mode  <- sub(".*_", "", tname) 
        return(df)
      }))
      df.pheno$pheno <- pheno
      return(df.pheno)
    })
  )
  return(combined_df)
}

diff_table <- get_diff_table(filedf)
```

## Table under "Sig" Mode
```{r}
library(DT)
diff_table_sig <- subset(diff_table, mode == "sig")
datatable(diff_table_sig, options = list(pageLength = 10))
```

## Table under "Reg" Mode
```{r}
diff_table_reg <- subset(diff_table, mode == "reg")
datatable(diff_table_reg, options = list(pageLength = 10))
```

# Plotting the Number of Significant Differential Genes
The provided function generates bar plots illustrating the number of significantly differentially expressed genes (**`dd.fdr < 0.1`**) for each tumor type across different phenotype contexts.
```{r}
plot_diff_number <- function(filedf, mode = c("all", "reg", "sig")){
  mode <- match.arg(mode)
  
  if (mode != "all"){
    filedf <- filedf[filedf$mode == mode, ]
  }
  
  pheno_all <- unique(filedf$phenotype)
  numlist <- list()
  
  for (p in pheno_all){
    p_txtfiles <- filedf[filedf$phenotype == p, ]
    numlist[[p]] <- list()
    for (t in 1:nrow(p_txtfiles)){
      txtf <- p_txtfiles[t, "filetxt"]
      tumor <- p_txtfiles[t, "tumor"]
      res <- read.table(txtf, header = TRUE)
      numlist[[p]][[tumor]] <- c(
        nrow(res[res$dd.fdr < 0.1,]),
        nrow(res) - nrow(res[res$dd.fdr < 0.1,])
      )
    }
  }
  
  par(mfrow = c(length(numlist),1), mar = c(2,4,1,0))
  for (phenotype in names(numlist)) {
    colors <- rainbow(length(numlist[[phenotype]]))
    plot(
      NULL,
      xlim = c(0.5, length(numlist[[phenotype]]) + 0.5),
      ylim = c(0, max(unlist(sapply(numlist[[phenotype]], `[`, 1)))),
      xlab = "Tumor Type",
      ylab = "No. Genes",
      main = paste("Context:", phenotype),
      xaxt = "n", bty = 'n'
    )
    grid()
    
    axis(
      1,
      at = 1:length(numlist[[phenotype]]),
      labels = names(numlist[[phenotype]]),
      las = 2
    )
    
    for (i in seq_along(numlist[[phenotype]])) {
      bar_height <- numlist[[phenotype]][[i]][1]
      rect(i - 0.4, 0, i + 0.4, bar_height, col = colors[i], border = NA)
    }
  }
}
```

## Visualization under "Reg" Mode
```{r}
plot_diff_number(filedf, mode = "reg")
```

## Visualization under "Sig" Mode
```{r}
plot_diff_number(filedf, mode = "sig")
```

# Plot specific genes
To visualize the data for specific genes, diffdriver has a plotting function:
```{r}
mut_path <- "/dartfs/rc/lab/S/Szhao/qiruiz/diffdriver/tumor_specific_input/HNSC/HNSC_mutations.txt"
mut <- read.table(mut_path, header = TRUE, sep = "\t")
head(mut)
```

```{r}
pheno_path <- "/dartfs/rc/lab/S/Szhao/qiruiz/diffdriver/cbioportal_download/Aneuploidy.Score.txt"
pheno <- read.table(pheno_path, header = TRUE, sep = "\t")
head(pheno)
```

```{r}
diffdriver::plot_mut(gene_name = "CDKN2A", mut= mut, pheno = pheno, totalnttype = 96, 
                     anno_dir =  "/dartfs/rc/lab/S/Szhao/qiruiz/diffdriver/temp/annodir96")
```
