---
title: "LUAD smoking-differential driver analysis"
date: "2026-07-12"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

This analysis runs **diffdriver** on TCGA lung adenocarcinoma (**LUAD**) to test, for a
panel of known LUAD driver genes, whether the strength of positive selection differs by
**smoking status**, adjusting for **age** and **sex** as covariates.

- **Phenotype:** smoking status, binary — ever-smoker (`tobacco_smoking_history_indicator` 2–5) = 1, never-smoker (= 1) = 0.
- **Covariates:** age (`age_at_initial_pathologic_diagnosis`), sex (`gender`).
- **Mutations:** TCGA LUAD MAF, SNVs only, keyed by 12-character patient barcode.
- **Model:** signature mode (`BMRmode = "signature"`, `k = 6`, 96 nucleotide types).
- **Samples analyzed (mutation ∩ phenotype ∩ covariates):** 461 (393 ever-smokers, 68 never-smokers).

# Source files

All inputs are on the Dartmouth Discovery cluster under
`/dartfs-hpc/rc/home/m/f0052zm/szhao_lab_share/`:

| Input | Path |
|---|---|
| Mutations (MAF) | `TCGA_data/Project_TCGA/SSM_data/analyses__2016_01_28/gdac.broadinstitute.org_LUAD-TP.MutSigNozzleReport2.0.Level_4.2016012800.0.0/LUAD-TP.final_analysis_set.maf` |
| Clinical | `diffDriver/data_clinical/GDC_download/LUAD/GDCdata/TCGA-LUAD/harmonized/Clinical/Clinical_Supplement/42bf5eb2-bc49-45be-b18a-290f712b006c/nationwidechildrens.org_clinical_patient_luad.txt` |
| Genes | `diffDriver/tumor_specific_input/LUAD/LUADgenes.txt` (19 genes) |

Outputs are written to `diffDriver/data_run_LUAD_known/`.

# Data preparation

The MAF is converted to the diffdriver mutation format (`Chromosome`, `Position`, `Ref`,
`Alt`, `SampleID`), keeping SNVs and using the patient barcode as `SampleID`. The clinical
file (3 header rows, data from row 4) is parsed into the phenotype (smoking) and covariate
(age, gender) tables. `prep_LUAD.R`:

```r
suppressMessages(library(data.table))
outdir <- ".../diffDriver/data_run_LUAD_known"
maf  <- ".../gdac.broadinstitute.org_LUAD-TP.MutSigNozzleReport2.0.Level_4.2016012800.0.0/LUAD-TP.final_analysis_set.maf"
clin <- ".../nationwidechildrens.org_clinical_patient_luad.txt"

## mutations: MAF -> diffdriver format (SNVs only)
m <- fread(maf, sep = "\t", header = TRUE,
           select = c("Chromosome", "Start_position", "Reference_Allele",
                      "Tumor_Seq_Allele1", "Tumor_Seq_Allele2",
                      "Tumor_Sample_Barcode", "Variant_Type"))
m <- m[Variant_Type == "SNP"]
m[, Alt := fifelse(Tumor_Seq_Allele2 != Reference_Allele, Tumor_Seq_Allele2, Tumor_Seq_Allele1)]
mut <- unique(data.table(Chromosome = m$Chromosome, Position = m$Start_position,
                         Ref = m$Reference_Allele, Alt = m$Alt,
                         SampleID = substr(m$Tumor_Sample_Barcode, 1, 12)))
fwrite(mut, file.path(outdir, "LUAD_mut.txt"), sep = "\t")

## clinical: phenotype (smoking, binary ever/never) + covariates (age, gender)
hdr <- as.character(unlist(fread(clin, sep = "\t", header = FALSE, nrows = 1)))
cl  <- fread(clin, sep = "\t", header = FALSE, skip = 3)
setnames(cl, hdr)
cl <- cl[, .(SampleID = bcr_patient_barcode, gender,
             smoke = as.character(tobacco_smoking_history_indicator),
             age = age_at_initial_pathologic_diagnosis)]
cl[, Smoking := NA_real_]
cl[smoke == "1", Smoking := 0]
cl[smoke %in% c("2", "3", "4", "5"), Smoking := 1]
pheno <- cl[!is.na(Smoking), .(SampleID, Smoking)]
fwrite(pheno, file.path(outdir, "LUAD_pheno.txt"), sep = "\t")
cov <- cl[, .(SampleID, Age = suppressWarnings(as.numeric(age)), Gender = gender)]
cov <- cov[!is.na(Age) & Gender %in% c("MALE", "FEMALE")]
fwrite(cov, file.path(outdir, "LUAD_covariates.txt"), sep = "\t")
```

This yields 174,053 SNVs across 545 samples; smoking known for 508 patients (433 ever /
75 never); complete age+gender for 503; **461 samples** shared across all three.

# Running diffdriver

diffdriver was run as a SLURM batch job on Discovery. `run_LUAD.R`:

```r
suppressMessages(library(diffdriver)); suppressMessages(library(data.table))
outdir <- ".../diffDriver/data_run_LUAD_known"
gene  <- read.table(".../tumor_specific_input/LUAD/LUADgenes.txt",
                    header = FALSE, stringsAsFactors = FALSE)[, 1]
mut   <- as.data.frame(fread(file.path(outdir, "LUAD_mut.txt")))
pheno <- as.data.frame(fread(file.path(outdir, "LUAD_pheno.txt")))
cov   <- as.data.frame(fread(file.path(outdir, "LUAD_covariates.txt")))
set.seed(0)
res <- diffdriver(gene = gene, mut = mut, pheno = pheno, covariates = cov,
                  anno_dir = "/dartfs/rc/lab/S/Szhao/library/diffdriver_anno/annodir96",
                  k = 6, totalnttype = 96, BMRmode = "signature",
                  output_dir = outdir, output_prefix = "LUAD_known")
```

`job_LUAD.sh` (submitted with `sbatch`):

```bash
#!/bin/bash
#SBATCH --job-name=diffdriver_LUAD
#SBATCH --output=.../data_run_LUAD_known/diffdriver_LUAD.out
#SBATCH --error=.../data_run_LUAD_known/diffdriver_LUAD.err
#SBATCH --time=24:00:00
#SBATCH --mem=64G
#SBATCH --cpus-per-task=8
set -euo pipefail
module load R
cd .../data_run_LUAD_known
Rscript run_LUAD.R
```

The job completed in ~1h44m and wrote `LUAD_known_Smoking_resdd.txt` (results table),
`LUAD_known_Smoking_resdd.Rd` (full objects), and `LUAD_known_BMRres.Rdata` (background
model).

# Results

`dd.p` / `dd.fdr` are the diffdriver covariate-adjusted p-value and FDR for
smoking-differential selection; `mut.E1`/`mut.E0` are the number of mutated samples among
ever- (`E1 = 393`) and never-smokers (`E0 = 68`).

```{r results}
library(DT)
res <- read.table("data/LUAD_known_Smoking_resdd.txt", header = TRUE)
res$gene <- rownames(res)
res <- res[order(res$dd.p),
           c("gene", "dd.p", "dd.fdr", "mut.E1", "mut.E0", "E1", "E0",
             "mlr.p", "fisher.p", "binom.p", "lr.p")]
datatable(res, rownames = FALSE,
          extensions = "Buttons",
          options = list(dom = "Bfrtip",
                         buttons = c("copy", "csv", "excel"),
                         pageLength = 20)) |>
  formatSignif(columns = c("dd.p", "dd.fdr", "mlr.p", "fisher.p", "binom.p", "lr.p"), digits = 3)
```

**EGFR** is the only gene significant after multiple-testing correction
(`dd.p = 0.0019`, `dd.fdr = 0.037`). This is consistent with the well-established biology
that EGFR-mutant lung adenocarcinoma is enriched in never-smokers. Notably, the simple
frequency-comparison methods do **not** flag EGFR (`fisher.p = 0.42`, `mlr.p = 0.20`),
because raw mutation frequencies are similar between groups (10.9% in ever- vs 14.7% in
never-smokers); diffdriver detects it because it accounts for the lower background mutation
rate in never-smokers, revealing stronger *selection* on EGFR there.

No other gene passes FDR 0.05 (CDKN2A is nominal at `dd.p = 0.036`), consistent with the
limited power from the small never-smoker group (68 of 461) given the ~85/15 smoking
imbalance in LUAD.

# Combined TCGA + OncoSG cohort (cBioPortal)

## Motivation

The TCGA-only analysis above had only **68 never-smokers** (an ~85/15 ever/never split),
which limits power to detect smoking-differential selection. To improve balance and power we
assembled a larger whole-exome LUAD cohort from [cBioPortal](https://www.cbioportal.org/),
adding the **OncoSG** Singapore/Asian cohort, which is strongly enriched for never-smokers.

## Source files

Data were pulled from the cBioPortal REST API (`https://www.cbioportal.org/api`). Two
**whole-exome-sequenced (WES)**, **GRCh37/hg19** lung adenocarcinoma studies were used
(WES filtering via each sample's gene panel; targeted-panel studies such as MSK-IMPACT were
excluded):

| Study id | Cohort | WES samples | Population |
|---|---|---|---|
| `luad_tcga_pan_can_atlas_2018` | TCGA | 566 | mostly White smokers |
| `luad_oncosg_2020` | OncoSG | 302 | Asian, never-smoker-enriched |

- **Mutations:** `POST /molecular-profiles/{study}_mutations/mutations/fetch`, restricted to
  single-nucleotide variants and de-duplicated.
- **Clinical:** age, sex, and ethnicity from the patient clinical data; smoking from
  `TOBACCO_SMOKING_HISTORY_INDICATOR` (TCGA; 1 = never, 2–5 = ever) and `SMOKING_STATUS`
  (OncoSG; Yes/No). Ethnicity harmonized to White / Black / Asian.
- Samples were de-duplicated to one per patient; the two studies share no patients.

The assembled inputs live on Discovery in `data_run_LUAD_known/`: `LUAD_cbio_mut.txt`
(183,619 SNVs, 863 samples), `LUAD_cbio_pheno.txt` (smoking), `LUAD_cbio_cov.txt`
(age, sex, cohort), and `LUAD_cbio_anc_pheno.txt` / `LUAD_cbio_anc_cov.txt` (ancestry run).
The same 19-gene LUAD panel and 96-nucleotide signature model were used.

## Sample characteristics

- **868 WES samples** total (863 with mutation data; 183,619 SNVs).
- **Smoking** (analysis 1): 774 samples with known status — **521 ever / 253 never**
  (67% smokers). By cohort: TCGA 427 ever / 73 never; OncoSG 111 ever / 187 never. This is
  ~4× more never-smokers than the TCGA-only run (68), and a far more balanced split.
- **Ethnicity:** TCGA is 388 White / 52 Black / 9 Asian; OncoSG is 302 Asian. So ancestry is
  almost perfectly confounded with cohort (Asian ⟺ OncoSG).
- **Ancestry** (analysis 2): 694 samples — **311 Asian / 388 White** (Black and
  unknown-ancestry samples excluded).

Two analyses were run on this cohort:

1. **Smoking** (ever = 1 / never = 0), covariates **age, sex, cohort** — cohort is included
   to absorb the TCGA-vs-OncoSG study/batch difference.
2. **Ancestry** (Asian = 1 / White = 0), covariates **age, sex** — cohort *cannot* be
   included here because it is collinear with the phenotype.

## Smoking results (combined cohort)

`mut.E1`/`mut.E0` are mutated-sample counts among ever- (`E1 = 521`) and never-smokers
(`E0 = 253`).

```{r cbio_smoking}
res <- read.table("data/LUAD_cbio_Smoking_resdd.txt", header = TRUE)
res$gene <- rownames(res)
res <- res[order(res$dd.p),
           c("gene", "dd.p", "dd.fdr", "mut.E1", "mut.E0", "E1", "E0",
             "mlr.p", "fisher.p", "binom.p", "lr.p")]
datatable(res, rownames = FALSE, extensions = "Buttons",
          options = list(dom = "Bfrtip", buttons = c("copy", "csv", "excel"), pageLength = 20)) |>
  formatSignif(columns = c("dd.p", "dd.fdr", "mlr.p", "fisher.p", "binom.p", "lr.p"), digits = 3)
```

With the larger, balanced cohort the analysis recovers the canonical LUAD smoking biology and
detects many more differential drivers than the TCGA-only run (which found only EGFR):

- **Smoking-associated (more selected in ever-smokers):** **KRAS** (`dd.fdr = 2e-4`), **NF1**
  (`4e-3`), **KEAP1** (`0.025`), **STK11** (`0.034`), and DZIP1L / LCE1F / ARID2
  (`~0.03–0.05`). TP53 is borderline (`dd.fdr = 0.059`).
- **Never-smoker-associated:** **EGFR** (`dd.fdr = 0.032`; 69/521 ever vs 79/253 never) —
  again enriched in never-smokers, consistent with the TCGA-only result and known biology.

This matches the well-established pattern that KRAS/STK11/KEAP1/NF1 mark smoking-driven LUAD
whereas EGFR marks never-smoker LUAD.

## Ancestry results (Asian vs White)

`mut.E1`/`mut.E0` are mutated-sample counts among Asian (`E1`) and White (`E0`).

```{r cbio_ancestry}
resA <- read.table("data/LUAD_cbio_anc_Asian_resdd.txt", header = TRUE)
resA$gene <- rownames(resA)
resA <- resA[order(resA$dd.p),
             c("gene", "dd.p", "dd.fdr", "mut.E1", "mut.E0", "E1", "E0",
               "mlr.p", "fisher.p", "binom.p", "lr.p")]
datatable(resA, rownames = FALSE, extensions = "Buttons",
          options = list(dom = "Bfrtip", buttons = c("copy", "csv", "excel"), pageLength = 20)) |>
  formatSignif(columns = c("dd.p", "dd.fdr", "mlr.p", "fisher.p", "binom.p", "lr.p"), digits = 3)
```

Many genes reach high significance: **EGFR** (`dd.p = 5e-11`, strongly enriched in Asian
LUAD), **KEAP1** (`3e-11`), **KRAS** (`4e-10`), NF1, SMARCA4, TP53, MGA, SETD2, STK11,
ARID1A, PTPRU, BRAF. EGFR is enriched in Asian; KRAS/KEAP1/STK11/NF1 in White.

## Interpretation and caveats

In this cohort **smoking, ancestry, and cohort are all entangled**: Asian ≈ never-smoker ≈
OncoSG, and White ≈ smoker ≈ TCGA. Consequently the two analyses largely trace the *same*
biological axis (EGFR at one end, KRAS/STK11/KEAP1 at the other).

- The **smoking** analysis is the more reliable of the two because it explicitly adjusts for
  **cohort**, removing the bulk of the TCGA-vs-OncoSG study/batch difference; its findings are
  consistent with the independent TCGA-only run and with established biology.
- The **ancestry** analysis **cannot** adjust for cohort (cohort is collinear with the
  Asian/White phenotype), so its results conflate genuine ancestry biology — the EGFR
  enrichment in Asian LUAD is real and well documented — with residual batch effects from the
  two studies' different sequencing/mutation-calling pipelines. Its p-values should therefore
  be read as an upper bound on ancestry-specific selection, not as batch-free estimates.
