---
title: "Pan-cancer immune-subtype driver analysis (299 driver genes)"
date: "2026-07-19"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

This is an expanded pan-cancer version of the immune-subtype driver analysis. For each of a
large set of driver genes, we pool tumor samples across the cancer types in which the gene is
recurrently point-mutated, and test whether the gene's selection differs by **immune subtype**
(Thorsson C1–C4, one-vs-rest), adjusting for age, sex, and tumor type of origin.

- **Genes:** the 299 driver genes from Bailey et al. 2018 (Cell), filtered to those with a
  usable recurrently-mutated cohort (below).
- **Phenotype:** immune subtype, one-vs-rest — C1 vs rest, C2 vs rest, C3 vs rest, C4 vs rest.
- **Covariates:** age, sex, and **cohort (tumor type)**. Sex is dropped for single-sex
  cohorts (e.g. UCEC/OV), and cohort is dropped for single-tumor cohorts.
- **Model:** signature mode (`BMRmode = "signature"`), with the number of mutational
  signatures scaled to cohort size: `k = 6` for ≤1000 samples, +1 per 200 samples above
  1000, capped at 20.

# Data sources

- **Mutations:** TCGA PanCancer Atlas (32 studies) from [cBioPortal](https://www.cbioportal.org/),
  **point mutations (SNVs) only**, **excluding hypermutators** (>1000 mutations/sample).
- **Immune subtype (C1–C6):** Thorsson et al. 2018 (Immunity) pan-immune table, matched by
  TCGA participant barcode.
- **Age, sex:** cBioPortal clinical data.

**Gene / cohort selection.** Starting from the 299 Bailey drivers, for each gene we counted
point mutations per tumor type and kept the tumor types with ≥5 point mutations; genes with a
total of >20 point mutations were retained (254 genes). **SKCM (melanoma) was excluded**
(its non-hypermutator subset gives a degenerate topic-model matrix). Each gene's **cohort** is
the union of its qualifying tumor types among the 27 tumor types with immune-subtype labels;
genes sharing the same tumor set were grouped so the background mutation rate is estimated
once per cohort. This yields **237 cohorts across 8,253 samples**, run as `237 × 4` = 948
diffdriver analyses (BMR reused across the 4 subtypes).

Results are available for **220 genes** (6 small low-mutation cohorts — EIF1AX, SOX17, PCBP1,
IDH2, VHL, ELF3 — could not be fit in signature mode and are omitted).

# Per-gene cohorts and mutation frequency

Each gene's pan-cancer cohort: the tumor types pooled, the total number of samples, the number
of mutated samples, and the resulting **mutation frequency** (mutated samples / cohort
samples). Sorted by mutation frequency.

```{r cohorts}
library(DT)
gc <- read.table("data/pancancer299_gene_cohorts.txt", header = TRUE, sep = "\t")
gc <- gc[order(-gc$mut_freq), ]
gc$mut_freq_pct <- round(100 * gc$mut_freq, 1)
datatable(gc[, c("gene", "n_tumor_types", "n_samples", "n_mutated", "mut_freq_pct", "tumors")],
          rownames = FALSE, filter = "top", extensions = "Buttons",
          colnames = c("gene", "n tumor types", "cohort samples", "mutated samples",
                       "mut freq (%)", "cohort (tumor types)"),
          options = list(dom = "Bfrtip", buttons = c("copy", "csv", "excel"), pageLength = 15))
```

# Results

For each gene we take `dd.p` from its cohort run — the diffdriver likelihood-ratio test
p-value (−2·(null − alt) log-likelihood, compared to χ² with 1 df). We then **recompute FDR
by Benjamini–Hochberg across all genes within each subtype** (`p.adjust(method = "fdr")`
applied separately to the ~220 p-values for C1, C2, C3, and C4). `alpha_e` is the subtype
coefficient — **positive = selected up in the subtype, negative = up in the rest**.
`mut_freq` is the cohort-level mutation frequency; `mut.E1`/`mut.E0` are mutation counts in
the subtype vs the rest.

```{r results}
res <- read.table("data/pancancer299_immune_aggregated.txt", header = TRUE, sep = "\t")
res$mut_freq_pct <- round(100 * res$mut_freq, 1)
res <- res[order(res$phenotype, res$dd.p),
           c("gene", "phenotype", "dd.p", "dd.fdr", "direction", "mut_freq_pct",
             "n_tumor_types", "n_samples", "mut.E1", "mut.E0", "tumors")]
datatable(res, rownames = FALSE, filter = "top", extensions = "Buttons",
          options = list(dom = "Bfrtip", buttons = c("copy", "csv", "excel"), pageLength = 15)) |>
  formatSignif(columns = c("dd.p", "dd.fdr"), digits = 3)
```

## Significant genes (dd.fdr < 0.1)

- **C2 vs rest (11):** TP53 (up in C2), KRAS (up in rest), JAK2, COL5A1, CASP8, **HLA-A**,
  PPP2R1A, **HLA-B**, EZH2, RB1, FAT1 — all up in C2 except KRAS.
- **C3 vs rest (7):** TP53, WHSC1/NSD2, FAT1, SMARCA4, BRCA1, LATS2, POLQ — all up in the
  rest.
- **C4 vs rest (2):** CIC (up in rest), **EGFR (up in C4)**.
- **C1 vs rest:** none pass FDR 0.1.

The dominant, most consistent signal is **TP53**, which flips direction — enriched in C2
("IFN-γ dominant") but depleted (up in rest) for C3 ("inflammatory") — consistent with the
immune subtypes tracking distinct oncogenic-driver backgrounds. **KRAS** is depleted in C2.
At the 0.1 threshold, C2 additionally picks up the antigen-presentation genes **HLA-A** and
**HLA-B**, whose selection tracks immune-subtype differences in immune evasion. Notably
**EGFR** is selected up in **C4** (lymphocyte-depleted), and **CIC** in the rest of C4. This
cohort-adjusted pan-cancer view (250 genes, 27 tumor types) extends the earlier
tumor-type-restricted immune-subtype analyses.

# Notes

- FDR is recomputed within each subtype over the genes tested, so values differ from the
  per-tumor-type immune-subtype pages.
- The full aggregated table (all 220 genes × 4 subtypes, with mutation frequency and cohort
  composition) is downloadable from the results table above.
