---
title: "Pan-cancer immune-subtype driver analysis"
date: "2026-07-14"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

The per-tumor-type immune-subtype analyses (see [Immune Subtypes](immunesubtypes.html)) run
diffdriver separately within each tumor type. This analysis instead performs a **pan-cancer**
test: for each driver gene, it pools samples across the tumor types where the gene is
recurrently mutated, and asks whether the gene's selection differs by immune subtype across
the whole pan-cancer cohort — adjusting for the tumor type of origin.

- **Phenotype:** immune subtype, each tested one-vs-rest — C1 vs rest, C2 vs rest, C3 vs rest,
  C4 vs rest.
- **Covariates:** age, sex, and **cohort (tumor type)** — the cohort term absorbs
  tumor-type-specific mutation-rate and composition differences.
- **Model:** signature mode (`BMRmode = "signature"`, `k = 6`, 96 nucleotide types).

# Gene selection and cohort construction

Starting from the per-tumor-type diffdriver outputs across **19 tumor types**
(BLCA, BRCA, CESC, CHOL, ESCA, GBM, HNSC, KICH, KIRC, KIRP, LIHC, LUAD, LUSC, PAAD, PRAD,
SARC, TGCT, UCEC, UCS — SKCM excluded, its runs had insufficient data), we counted, per gene,
the number of mutated samples in each tumor type. For each gene we then formed a **pan-cancer
cohort = the union of all tumor types in which the gene has at least 5 mutations**.

We analyzed the genes whose cohort spans **at least two tumor types** — **43 genes**. Genes
sharing the same tumor-type set were grouped into a common cohort so the background mutation
rate is estimated once per cohort (36 distinct cohorts). Each cohort × subtype combination is
one diffdriver run (36 × 4 = 144 runs), submitted as a SLURM array on Discovery.

```{r cohorts}
library(DT)
pc <- read.table("data/pancancer_gene_cohorts.txt", header = TRUE, sep = "\t")
pc <- pc[pc$n_tumor_types >= 2, ]
pc <- pc[order(-pc$samples_with_mut), c("gene", "n_tumor_types", "total_samples",
                                        "samples_with_mut", "tumor_types")]
datatable(pc, rownames = FALSE, extensions = "Buttons",
          options = list(dom = "Bfrtip", buttons = c("copy", "csv", "excel"), pageLength = 10),
          colnames = c("gene", "n tumor types", "total samples", "samples w/ mut",
                       "tumor types (mutation count)"))
```

Covariates are assembled from three sources on Discovery, keyed on the 12-character TCGA
patient barcode: mutations from each `<TUMOR>_mutations.txt`; immune subtype (C1–C6, one-hot)
and tumor label from `<TUMOR>_6immunesubtype.txt`; age and sex from the cBioPortal
`combined_study_clinical_data.tsv`. Across the 18 contributing tumor types this gives 4,972
samples with complete age and sex.

# Results

For each gene we take `dd.p` from its pan-cancer cohort run, then **recompute FDR across the
gene set within each subtype** (FDR depends on the number of genes tested). `alpha_e` is the
subtype coefficient: **positive = selected up in the subtype, negative = up in the rest**.
`mut.E1`/`mut.E0` are mutation counts in the subtype vs the rest.

```{r results}
res <- read.table("data/pancancer_immune_aggregated.txt", header = TRUE, sep = "\t")
res <- res[order(res$phenotype, res$dd.p),
           c("gene", "phenotype", "dd.p", "dd.fdr", "alpha_e", "direction",
             "mut.E1", "mut.E0", "E1", "E0", "tumors")]
datatable(res, rownames = FALSE, extensions = "Buttons", filter = "top",
          options = list(dom = "Bfrtip", buttons = c("copy", "csv", "excel"), pageLength = 15)) |>
  formatSignif(columns = c("dd.p", "dd.fdr", "alpha_e"), digits = 3)
```

## Significant genes (dd.fdr < 0.05)

- **C2 vs rest (6 genes):** KRAS (up in rest), TP53 (up in C2), CTNNB1 (up in rest),
  FOXA1 (up in rest), CASP8 (up in C2), HRAS (up in C2).
- **C3 vs rest (2 genes):** TP53 (up in rest), KRAS (up in C3).
- **C1 and C4:** no genes pass FDR 0.05.

The dominant, most consistent signals are **TP53 and KRAS**, which flip direction between
subtypes — TP53 is selected *up* in C2 but *down* in C3, and KRAS the reverse. This is
consistent with the immune subtypes (C2 "IFN-γ dominant", C3 "inflammatory") tracking distinct
oncogenic-driver backgrounds, since TP53 and KRAS status are both known to shape the tumor
immune microenvironment. FOXA1, CTNNB1, CASP8, and HRAS contribute subtype-specific selection
within their respective tumor cohorts.

# Notes

- **42 of 43 genes** are reported. **SPOP** could not be fit — a numerical failure in the EM
  optimization within its PRAD+UCEC+UCS cohort.
- FDR is recomputed within each subtype over the genes tested, so the values here differ from
  the per-tumor-type FDRs on the [Immune Subtypes](immunesubtypes.html) page.
- The full aggregated table (all 42 genes × 4 subtypes, with mutation counts and cohort
  composition) is downloadable from the results table above.
